Exploring the composition of lithic assemblages in Mesolithic south-eastern Norway

### 2021-05-13 analysis/paper/paper.Rmd
Curation index on both flint and non-flint. Highlighted sites of interest

Previous studies of lithic artefacts in Norway have typically had a focus on chronological trends associated with formal tool types [e.g. @helskog1976; @reitan2016], involved refit-studies of the entirety of individual assemblages [@skar1986], or been concerned with technological processes associated with certain sub-categories of the site inventories, such as blades or axes [e.g. @damlien2016; @berg-hansen2017; @solheim2020]. Some studies have involved in-depth analyses of handful of sites in terms of intrasite spatial organisation [@nærøy2018], and technological organisation [@eigeland2015; @mansrud2016;]. There are also some more comprehensive evaluations of a larger number of sites with associated lithic assemblages to get at potential site types and related mobility patterns [@breivik2016; @viken2018c; @breivik2020]. However, the narratively driven analyses of these data makes the weighting of different variables unclear, and these studies draw their conclusions based on an arguably over-constraining predefined trichotomy of different site types. Finally, only a single study concerned with eight Middle-Mesolithic sites has employed a multivariate quantitative framework to help structure such a treatment [@solheim2013a]. These sites are also included here. In sum, therefore, previous studies are typically either limited to a small number of sites, to a subset of the inventories, or to more subjectively and narratively driven methods that makes it difficult, if not impossible, to consistently structure and balance the comparison of a larger number of sites with associated assemblages holding lithic artefacts numbering in the thousands.

@@

As the title of the paper suggest, the data has been analysed from an exploratory perspective, meaning that a wide range of methods has been applied. Only parts of this process can sensibly be reported here. The data and R script used for the analysis is freely available as a research compendium following @marwick , allowing readers to explore the data and the analytical choices made [cf. @Marwick2017]. However, this kind of data dredging does arguably not constitute a good foundation for the pursuit of causal relationships, as this process is most certainly impacted by and selectively reported upon, following my own biases. It can, however, result in the identification of interesting relationships that can be subjected to more rigorous testing using different data and approaches in subsequent studies.

@@

However, following not least from the ambiguous definition first put forward by @binford1979, the theoretical definition of curation, its archaeological correlates, and behavioural implications have been widely discussed, and no single definition has ever been reached [e.g. @bamforth1986; @shott1996; @surovell2009, 9--13]. The continuum between curated and expedient technology has for example been related to dimensions such as land-use and mobility strategies, as well as raw-material quality and availability [e.g. @parry1987; @kuhn1992; @andrefsky1994; @smith2015; @clark2017]. Still, that the distinction can offer a useful analytical framework if clearly operationalised seems more or less agreed upon, and some dimensions of the concept are generally accepted. For example, although precisely how it is measured may vary, the empirical correspondent to a curated technological organisation is typically defined by high degrees of retouch, as this is commonly seen as a means of realising the potential utility of a tool --- or extending its use-life --- by the repeated rejuvenation and modification of edges [e.g. @bamforth1986; @dibble1995; @shott2005]. Furthermore, one concrete operationalisation of the terms have been forwarded by Barton and collegues [e.g. -@barton1999; @barton2014; @clark2017], who through a series of studies have shown that the relationship between volumetric density of lithics and relative frequency of retouched artefacts in lithic assemblages provide consistent results across a wide range of chronological and cultural context, and is a measure taken to reflect degree of curation. This is in turn mainly to follow from the accumulated nature of the land-use and mobility patterns associated with the sites [@barton2014]. In this model, higher degree of mobility would mean a higher dependency on the artefacts and the material people could bring with them, and dimensions such as weight, reliabilty, repairability, and the degree to which artefacts could be manipulated to fulfill a wide range of tasks are therefore assumed to have been factors of concern. From this it follows that the empirical expectation for short-term camps is a curated technological organisation with higher relative frequency of retouched artefacts, and a lower overall density of lithics [@clark2017]. More time spent in a single location, on the other hand, is assumed to lead to better control of raw-material availability and to allow for its accumulation. This should in turn lead to an expedient technological organisation with reduced necessity for the conservation of lithics and extensive use of retouch. The empirical expectation for lower degree of mobility is therefore relatively high density of lithics, a low relative frequency of retouched artefacts, as well as a higher number of cores and unretouched flakes and blades [@clark2017].

@@

```{r eigenvalues, fig.cap = "Table showing the eigenvalues of the components."}
# Create a table showing the eigenvalues of the components
round(as_tibble(get_eig(pcalithics)), 2) %>% 
  rename("Eigenvalue" = eigenvalue, "Variance (%)" = variance.percent, 
         "Cumulative variance (%)" = cumulative.variance.percent) %>% 
  rownames_to_column(var = "Component") %>% 
gt(rowname_col = "row", auto_align = FALSE)
```

@@

@@

# Curation index plot simply using mean site age (not used)
w3 <- ggplot(indexdata, aes(x = mean_date, y = index)) + 
  geom_point() +
  labs(title = "C",
       x = "Mean date cal. BCE",
       y = "Curation index") +
  scale_x_reverse() + 
  geom_smooth(method = "loess", col = "red") +
  theme_classic()  + 
  ggrepel::geom_label_repel(aes(label = site),box.padding   = 0.35, 
                  point.padding = 0.5,
                  segment.color = 'grey50') 
#w3
```

```{r simage, cache = TRUE, fig.cap = "Temporal variation in the curation index. The temporal uncertainty is handled by means of a simulation approach, where the site ages are drawn from their associated probability distributions given in Figure 1B. A LOESS curve has been fit to the distribution for each of the 1000 simulation runs."}

@@

# Create data frame to hold simulation results.
results <- data.frame(matrix(NA, ncol = nsim, nrow = nrow(sites))) 
# Insert the curation index value for each site
results[,"index"] <- agedata %>% 
  group_by(site) %>% 
  slice(1) %>% 
  ungroup() %>% 
  select(index) %>% 
  c()

@@

# Call to plot, grouping loess fits by simulation runs
ggplot(resultslong, aes(x = age, y = index)) + 
  geom_smooth(aes(group = sim), method = "loess", fill = "grey", col = "black", alpha = 0.05, se = TRUE) +
  geom_point(size = 0.5, alpha = 0.005) +
  xlab("Cal. BCE") +
  ylab("Curation index") +
  theme_classic()

@@

Assuming for now that the curation index captures degree of mobility, the present data indicates two major modes with a generally higher degree of mobility until some time before 8000 BCE. This then drops and stabilises from around 7000 BCE for the rest of the Mesolithic. The variation in degree of curation, which, as a note, is at it highest between c. 8000 and 7000 BCE, could indicate that the sites were associated with a more varied mobility pattern after around 8000 BCE. The five sites that have values on the curation index below -0.25, could in this perspective have predominantly functioned as base camps within a logistic settlement pattern. That these assemblages reflect stays of a longer duration was suggested for all five sites in the original reports, although on different and differing grounds than the curation measures employed here. Another case worth commenting on is sites such as Langangen Vestgård 1, which on the grounds of a large number of artefacts and the possibile presence of a dwelling structure was argued to reflect a more stable site location in the original report. However, the site could reflect the aggregation of stays which predominantly have been of shorter duration, while the possible dwelling structure, if taken as an indication of lower mobility, could represent a remnant from one or a few longer stays that constitute a smaller fraction of the use-life of the site as a whole. It is also interesting to note that some of the chronological patterns indicated in the CA does not appear to be related to the curation measures. This for example pertains to the apparent importance of the axe production sites. It has recently been suggested that the end of the Mesolithic sees a drop in population numbers and a subsequent increase in mobility as migrating populations arrived from southern Scandinavia. Whether the apparent absence of this pattern in the present data is a result of a strained sample size or a weaker or non-existent signal from any changes in mobility patterns in the later Mesolithic remains an open question. As it stands, however, the main hypothesis generated by the present data would be that settlement patterns in the earliest parts of the Mesolithic were characterised by relatively high degrees of mobility, which then drop before leveling off at around 7000 BCE and remain relatively stable throughout the rest of the period. Given the parallel increase in variation this would seem to correlate well with a transition from a residential to a logistical settlement system.

As for the measures associated with curation and mobility patterns, relative frequency of secondarily worked lithics and volumentric density of lithics correspond most clearly to patterns previously proposed in the literature. However, as these were identified by employing PCA and identifying what variables most clearly correspond to a temporal dimension, there is a clear danger of circular reasoning here. Having established that these measures appears to capture a behavioural dimension of interest, this can, however, now be tested more rigorously by employing site data from a wider region. these results do speak to the cross-cultural and cross-temporal applicability of the measure, and consequently the potential it holds for comparative studies. The findings indicated by the WABI, as well as by the negative correlation between relative frequency of primarily and secondarily worked lithics, do align more clearly with previous suggestions concerning the overall mobility patterns in Mesolithic Norway. However, the question then becomes: What is the significance of the negatively correlated variables of relative frequency of cores and chips? Given that these are more or less orthogonal to the WABI variables on the first and second component, this would indicate that if one of these variable pairs is taken to reflect the scale between expedient and curated technologies, it should follow that the other variable pair is not correlated with these. As the entire temporal range of the sites is spread out between the extremes of the chip and core variables, one possibility is that these are indicative of site types that are temporally ubiquitous.

>>>>

Previous studies of lithic artefacts in Norway have typically had a focus on chronological trends associated with formal tool types [e.g. @helskog1976; @reitan2016], involved refit-studies of the entirety of individual assemblages [@skar1986], or been concerned with technological processes associated with certain sub-categories of the site inventories, such as blades or axes [e.g. @damlien2016; @berg-hansen2017; @solheim2020]. Some studies have involved in-depth analyses of handful of sites in terms of intrasite spatial organisation [@nærøy2018], and technological organisation [@eigeland2015; @mansrud2016;]. There are also some more comprehensive evaluations of a larger number of sites with associated lithic assemblages to get at potential site types and related mobility patterns [@breivik2016; @viken2018 ; @breivik2020]. However, the narratively driven analyses of these data makes the weighting of different variables unclear, and these studies draw their conclusions based on an arguably over-constraining predefined trichotomy of different site types. Finally, only a single study concerned with eight Middle-Mesolithic sites has employed a multivariate quantitative framework to help structure such a treatment [@solheim2013a]. These sites are also included here. In sum, therefore, previous studies are typically either limited to a small number of sites, to a subset of the inventories, or to more subjectively and narratively driven methods that makes it difficult, if not impossible, to consistently balance and structure the comparison of a larger number of sites with associated assemblages holding lithic artefacts numbering in the thousands.

@@

As the title of the paper suggest, the data has been analysed from an exploratory perspective, meaning that a wide range of methods has been applied, and various combinations and transformations of variables has been explored. Only parts of this process can sensibly be reported. However, this kind of data dredging does arguably not constitute a good foundation for the pursuit of causal relationships, as this process is most certainly selectively reported upon and heavily impacted by my own biases. It can, however, result in the identification of interesting relationships that can be subjected to more rigorous testing using different data and approaches in subsequent studies. Furthermore, the employed data and R script is freely available as a research compendium following @marwick, allowing readers to explore and scrutinise the data and the final analytical choices made [cf. @Marwick2017].

@@

However, following not least from the ambiguous definition first put forward by @binford1979, the theoretical definition of curation, its archaeological correlates, and behavioural implications have been widely discussed, and no single definition has ever been reached [e.g. @bamforth1986; @shott1996; @surovell2009, 9--13]. The continuum between curated and expedient technology has for example been related to dimensions such as land-use and mobility strategies, as well as raw-material quality and availability [e.g. @parry1987; @kuhn1992; @andrefsky1994; @smith2015; @clark2017]. Still, that the distinction can offer a useful analytical point of departure if clearly and explicitly operationalised seems more or less agreed upon, and some dimensions of the concept are generally accepted. For example, although precisely how it is measured may vary, the empirical correspondent to a curated technological organisation is typically defined by high degrees of retouch, as this is commonly seen as a means of realising the potential utility of a tool --- or extending its use-life --- by the repeated rejuvenation and modification of edges [e.g. @bamforth1986; @dibble1995; @shott2005]. Furthermore, one concrete operationalisation of the terms have been forwarded by Barton and collegues [e.g. -@barton1999; @barton2014; @clark2017], who through a series of studies have shown that the relationship between volumetric density of lithics and relative frequency of retouched artefacts in lithic assemblages provide consistent results across a wide range of chronological and cultural context, and is a measure taken to reflect degree of curation. This is in turn mainly to follow from the accumulated nature of land-use and mobility patterns associated with the sites [@barton2014]. In this model, higher degree of mobility would mean a higher dependency on the artefacts and the material people could bring with them, and dimensions such as weight, reliabilty, repairability, and the degree to which artefacts could be manipulated to fulfill a wide range of tasks are therefore assumed to have been factors of concern. From this it follows that the empirical expectation for short-term camps is a curated technological organisation with higher relative frequency of retouched artefacts, and a lower overall density of lithics [@clark2017]. More time spent in a single location, on the other hand, is assumed to lead to better control of raw-material availability and to allow for its accumulation. This should in turn lead to an expedient technological organisation with reduced necessity for the conservation of lithics and extensive use of retouch. The empirical expectation for lower degree of mobility is therefore relatively high density of lithics, a low relative frequency of retouched artefacts, as well as a higher number of cores and unretouched flakes and blades [@clark2017].

@@

@@

minmaxflint <- flint2 %>% 
   rownames_to_column(var = "site") %>% 
  mutate(dens =  ((flint_dens - min(flint_dens)) / (max(flint_dens) - min(flint_dens))),
         secondary = ((secondary_freq - min(secondary_freq)) / (max(secondary_freq) - min(secondary_freq))))

@@

indexdataflint <- minmaxflint %>% 
  rowwise() %>% 
  mutate(index_flint = mean(c(dens * -1, secondary)))

indexdata$index_flint <- indexdataflint$index_flint

# Curation index plot simply using mean site age (not used)
w3 <- ggplot(indexdata, aes(x = mean_date, y = index)) + 
  geom_point() +
  labs(title = "C",
       x = "Mean date cal. BCE",
       y = "Curation index") +
  scale_x_reverse() + 
  geom_smooth(method = "loess", col = "red") +
  theme_classic()  + 
  ggrepel::geom_label_repel(aes(label = site),box.padding   = 0.35, 
                  point.padding = 0.5,
                  segment.color = 'grey50')

w4 <- ggplot(indexdata, aes(x = mean_date, y = index_flint)) + 
  geom_point() +
  labs(title = "C",
       x = "Mean date cal. BCE",
       y = "Curation index") +
  scale_x_reverse() + 
  geom_smooth(method = "loess", col = "red") +
  theme_classic()  + 
  ggrepel::geom_label_repel(aes(label = site),box.padding   = 0.35, 
                  point.padding = 0.5,
                  segment.color = 'grey50') 
#w3
```

```{r simage, cache = TRUE, fig.cap = "Temporal variation in the curation index. The temporal uncertainty is handled by means of a simulation approach, where the site ages are drawn from their associated probability distributions given in Figure 1B. A LOESS curve has been fit to the distribution for each of the 1000 simulation runs. Each simulation is plotted with some transparency. Sites mentioned in the text are coloured.", fig.height = 10, fig.width = 12}

@@

# Create data frame to hold simulation results.
results <- data.frame(matrix(NA, ncol = nsim, nrow = nrow(sites))) 
# Insert the curation index value for each site
index_values <- agedata %>% 
  group_by(site) %>% 
  slice(1) %>% 
  ungroup() %>% 
  select(index, index_flint) %>% 
  c()

results[,"index"] <- index_values$index
results[,"index_flint"] <- index_values$index_flint

@@

# Create column to hold whether site is to be coloured specifically in the plots
# (pertains to sites discussed in the text)
sitelabs <- c("Hovland 3", "Langangen Vestgård 1", "Hegna vest 4",
              "Vallermyrene 4", "Gunnarsrød 6b", "Prestemoen 1")

resultslong <- resultslong %>% 
  mutate(site_col = ifelse(site %in% sitelabs, site, "other"))

# Call to plot, grouping loess fits by simulation runs
i1 <- ggplot(resultslong, aes(x = age, y = index, colour = site_col)) + 
  geom_smooth(aes(group = sim), method = "loess", fill = "grey", col = "black",
              alpha = 0.05, se = TRUE) +
  geom_point(size = 0.5, alpha = 0.01) +
  scale_color_manual(name = NULL, breaks = c("Langangen Vestgård 1", 
                                             "Hegna vest 4","Gunnarsrød 6b",
                                             "Prestemoen 1",  "Hovland 3",
                                             "Vallermyrene 4"),
                    values = c("Langangen Vestgård 1" = "red",
                               "Hegna vest 4" = "blue",
                               "Gunnarsrød 6b" = "violet",
                               "Prestemoen 1" = "darkorange",
                               "Hovland 3" = "green",
                               "Vallermyrene 4"  = "cyan1", 
                               "other" = "black"),
                     guide = guide_legend(override.aes = list(size = 3,
                                                              alpha = 1))) +
  scale_x_continuous(breaks = c(seq(-9000,-4000, 1000))) +
  labs(title = "A", x = "Cal. BCE", y = "Curation index (all lithics)") +
  theme_classic() +
  theme(legend.position = "none")

i2 <- ggplot(resultslong, aes(x = age, y = index_flint, colour = site_col)) + 
  geom_smooth(aes(group = sim), method = "loess", fill = "grey", col = "black",
              alpha = 0.05, se = TRUE) +
  geom_point(size = 0.5, alpha = 0.01) +
  scale_color_manual(name = NULL, breaks = c("Langangen Vestgård 1", 
                                             "Hegna vest 4","Gunnarsrød 6b",
                                             "Prestemoen 1",  "Hovland 3",
                                             "Vallermyrene 4"),
                    values = c("Langangen Vestgård 1" = "red",
                               "Hegna vest 4" = "blue",
                               "Gunnarsrød 6b" = "violet",
                               "Prestemoen 1" = "darkorange",
                               "Hovland 3" = "green",
                               "Vallermyrene 4"  = "cyan1", 
                               "other" = "black"),
                     guide = guide_legend(override.aes = list(size = 3,
                                                              alpha = 1))) +
  scale_x_continuous(breaks = c(seq(-9000,-4000, 1000))) +
  labs(title = "A", x = "Cal. BCE", y = "Curation index (flint)") +
  theme_classic() +
  theme(legend.position = "bottom")

  i1 /
  i2
```

@@

Assuming for now that the curation index captures degree of mobility, the present data indicates two major modes with a generally higher degree of mobility until some time before 8000 BCE. This then drops and stabilises from around 7000 BCE for the rest of the Mesolithic. This pattern is evident in both the flint data and when all lithics are treated in aggregate. The variation in degree of curation when all lithics are considered could indicate that the sites were associated with a more varied mobility pattern after around 8000 BCE. The five sites that have values on the curation index below c. -0.25 could in this perspective have predominantly functioned as base camps within a logistic settlement pattern. That these assemblages reflect stays of a longer duration was suggested for all five sites in the original reports [@solheim2013; @carrasco2014; @eigeland2014b; @persson2014; @eigeland2017], with the exception of for Vallermyrene 4, which was argued to be a specialised axe production site, not necessarily associated with lower degrees of mobility. This highlights a possible issue pertaining to raw-material variability, as the coarse non-flint material used for the production of axes generally results in a relatively large amount of waste per produced tool, possibly skewing the curation index when compared to assemblages dominated by flint. However, even if this renders it more problematic to directly align the values on the curation index with a continuum between a residental and logistic settlement system, the variation on the index after around 8000 BCE is still evident. Another case worth commenting on is Langangen Vestgård 1, which, on the grounds of a large number of artefacts and the possibile presence of a dwelling structure was argued to reflect a more stable site location in the original report [@melvold2014]. However, the relatively high value on the curation index could mean that Langangen Vestgård 1 reflects the aggregation of stays which predominantly have been of a relatively shorter duration, while the possible dwelling structure, if taken as an indication of lower mobility, could represent a remnant from one or a few longer stays that constitute a smaller fraction of the use-life of the site as a whole.

While it is possible the is to consider these patterns at a population level. From this perspective it seems clear that there is a marked drop in curation from just before 8000 BCE, which stabilises around 7000 BCE. This would correspond well with chronological where the end of the Early Mesolithic, or flake axe phase, is set to last until c. 8200 BCE. The following phase linked to the production of microliths is in this framework set to last to around 7000 BCE, signifying the start of the production of the non-flint ground-pecked or round butted adzes. Later chronological transitions suggested at c. 5600 and c. 4500 BCE, are not reflected in fluctuations in the curation index. While this can be traced in the CA with the reduced importance of the non-flint categories, this could indicate that these changes are not related mobility patterns.

It is also interesting to note that some of the chronological patterns indicated in the CA does not appear to be related to the curation measures. This for example pertains to the apparent importance of the axe production sites. It has recently been suggested that the end of the Mesolithic sees a drop in population numbers and a subsequent increase in mobility as migrating populations arrived from southern Scandinavia. It is an open question whether the apparent absence of this pattern in the present data is a result of a strained sample size, or a weaker or non-existent signal from any changes in mobility patterns in the later Mesolithic. As it stands, however, the main hypothesis generated by the present data would be that settlement patterns in the earliest parts of the Mesolithic were characterised by relatively high degrees of mobility, which then drop before leveling off at around 7000 BCE and remain relatively stable throughout the rest of the period. Even though the relative increase in Given the parallel increase in variation this would seem to correlate well with a transition from a residential to a logistical settlement system.

As for the measures associated with curation and mobility patterns, relative frequency of secondarily worked lithics and volumentric density of lithics correspond most clearly to patterns previously proposed in the literature. However, as these were identified by employing PCA and identifying what variables most clearly correspond to a temporal dimension, there is a danger of circular reasoning. Having established that these measures appears to capture a behavioural dimension of interest, this can, however, now be tested more rigorously by employing site data from a wider region. these results do speak to the cross-cultural and cross-temporal applicability of the measure, and consequently the potential it holds for comparative studies. The findings indicated by the WABI, as well as by the negative correlation between relative frequency of primarily and secondarily worked lithics, do align more clearly with previous suggestions concerning the overall mobility patterns in Mesolithic Norway. However, the question then becomes: What is the significance of the negatively correlated variables of relative frequency of cores and chips? Given that these are more or less orthogonal to the WABI variables on the first and second component, this would indicate that if one of these variable pairs is taken to reflect the scale between expedient and curated technologies, it should follow that the other variable pair is not correlated with these. As the entire temporal range of the sites is spread out between the extremes of the chip and core variables, one possibility is that these are indicative of site types that are temporally ubiquitous.
